Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-11-19 paper/ifo-complexAudio.tex
full read-through before submission

% Optical microphone
% aim of optical microphone: play back injected sounds
% some editing in response to pnp
In this section we explore how this demonstration can be used to teach a selection of signal processing techniques.
The apparatus used in Sections~\ref{sec:ifo} to~\ref{sec:viterbi_wandering} affords the opportunity to explore a range of signal processing techniques and filters of interest to undergraduate students and provide a foundations for many technical applications~\cite{DigitalProcingOfSpeechSignals:1978}. 
We provide a series of example activities using complex audio (such as music and speech) as a natural successor to constant and wandering tones. 
%A natural successor to constant and wandering tones is complex audio, such as music and speech. 
As complex audio signals are not quasi-monochromatic, unlike continuous gravitational wave signals, the Viterbi algorithm as implemented in Section~\ref{sec:viterbi_wandering} is not directly applicable here. Instead, in this section, we use the interferometer as an \emph{optical microphone} to detect and play back complex audio signals (i.e.\ using light to capture sound).
In doing so, we experiment with a hierarchy of passive filters which suppress noise yet make no assumption about the specific form of the signal, unlike the Fourier-based maximum likelihood matched filter which is tuned to the sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood}.
A vast literature exists on signal processing with speech, however see Ref.~\cite{DigitalProcingOfSpeechSignals:1978} for introductory material including Fourier analysis and digital representations.
Optical microphone devices have precedence in the laser microphones~\cite{laser_microphone} which are (or were historically) used in the defence industry and operate on a variety of related principles.

%An introduction to ..... can be found in Ref.~\cite{}, including 
%A review of digital processing techniques for speech signals can be found 
%It is also analogous to the detection of unmodelled gravitational-wave signals, e.g. broadband bursts from supernovae~\cite{}\han{to do - add citation}. 
%It is also analogous to the detection of unmodelled gravitational wave signals, e.g. broadband bursts from supernovae. [refs; my last sentence can be improved]

This section is indirectly related to the advanced signal processing techniques required to extract gravitational-wave signals.
Here, the optical microphone serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories.
We describe the additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and initial results in Section~\ref{sec:initialResultsOpMic}. 
A selection of filter and speech enhancement example activities using the optical microphone are outlined in Section~\ref{sec:filters}.

@@

Audible frequencies are in the range of $\sim 20\,{\rm Hz}$--$20\,{\rm kHz}$. Speech intelligibility requires frequencies up to $3\,{\rm kHz}$ and music requires up to and beyond $8\,{\rm kHz}$~\cite{speech_intelligibility}. Therefore the optical microphone requires a sample rate of at least $16\,{\rm kHz}$ to capture both speech and music. This cannot be achieved with the webcam used in Sections~\ref{sec:single_tone} and ~\ref{sec:viterbi_wandering} as it has a sampling rate of $30\,{\rm Hz}$.
To overcome this issue, we use a photodiode\footnote{A photodiode is an electrical component that acts as a regular diode when no light is incident on it, blocking any current flow in the reverse direction. As the intensity of incident light rises, it becomes increasingly conductive in the reverse direction.} at the output of the interferometer to achieve a sampling rate of $16\,{\rm kHz}$.

We place a OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp which together make a photo-detector that produces a voltage dependent on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-centre position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}, again chosen arbitrarily.
The photodiode is mounted on a cloth screen re-purposed from the dismantled commercial speaker, with the electrical leads connected underneath. 
The voltage signal from the photo-detector is captured by a MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi (Pi) Model 3 v1.2~\cite{RaspberryPi:online}, which provides a convenient means to record the photodiode data.
Together, the circuit samples the signal at $\sim 16\,{\rm kHz}$. Resources for using the Pi and photodiode are described in Appendix~\ref{app:circuit_diagram}.

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing\footnote{Folding of frequencies greater than half the sampling rate.} into the detected range. We include an anti-aliasing Sallen-Key filter~\cite{sallen_key_filter} tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$, before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the intensity of the voltage signal and avoid saturating the ADC.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

\begin{figure*}
	\includegraphics[width=0.44\textwidth]{figures/ifo_schematic_photodiode_edit.pdf}
	\includegraphics[width=0.52\textwidth]{figures/setup_pic2.pdf}
	\caption{
Left: Michelson interferometer schematic. 
The schematic is identica.l to Figure~\ref{fig:interference_pattern} except that the data is now recorded using a photodiode and Raspberry Pi instead of a webcam. 
Right: photograph of the optical microphone. 
The Michelson interferometer is shown on the left and the circuit and Raspberry Pi on the right. 
In the main image, the photodiode is placed behind the cloth screen. 
The inset shows a face-on view of the photodiode with the covering screen removed. 
See Appendix~\ref{app:circuit_diagram} for a circuit diagram and detailed photograph of the circuit.
}
	\label{fig:ifo_schematic_podo}
\end{figure*}

@@

%The source audio was played through the speaker with the optical microphone recording for the full duration of the audio (around a minute long for most sources). 
We test the optical microphone with a variety of recordings including speech from different people and music ranging from simple melodies and rhythms to songs. During recordings, care is taken to minimise activity around the demonstration to reduce the amount of environmental noise coupling into the interferometer. The time series data is then directly converted to a .wav file and played as an audio recording using the \textbf{io.wavfile.write} function. When processing the results, we restrict our analysis to only the first $10\,{\rm s}$ of each observation (for efficiency), and only plot the first second here.

@@

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/psd_butterworth_14_6.pdf}
	\caption{\label{fig:psd_noise}
Power spectral density (PSD) of background noise (top) from the optical microphone (with the speaker off), and the PSD after applying a Butterworth bandpass filter (bottom) between the two frequencies marked with red, dashed lines. 
We see strong power from the $50\,{\rm Hz}$ mains hum and its harmonics (most likely from the photodiode circuit and the room’s lighting and cooling). Otherwise the PSD is fairly white. 
After filtering we see strong attenuation of all frequencies outside the band, but little change to the harmonics within the band.
}
\end{figure}

@@

Here, we describe several filters used to remove the $50\,{\rm Hz}$ mains hum and its harmonics, but also to ultimately improve the speech intelligibility of the recording (to be able to understand the words). We start with some naive approaches and then move to traditional signal processing filters like band-passing and cascaded notches. We finish by combining these with advanced statistical techniques like a Wiener filter and the logMMSE estimator. As an illustrative example, we test all of these filters on the same $1\,{\rm s}$ long speech recording.

\begin{figure*}
	\begin{center}
    %% trim order is <left> <lower> <right> <upper>
    % height=0.5\textheight,trim={1cm 4.5cm 0.4cm 3.5cm},clip
	\includegraphics[width=.85\textwidth, trim={1cm 3cm 1cm 1cm}]{figures/notch_and_wiener_superplot_v2.pdf}
	\end{center}
	\caption{\label{fig:BackgroundNotchWienerCombined}
	Time series (left column) and frequency spectrum (right column) results for the background subtraction, notch, and Wiener filters.
	The first and second rows show the original speech recording and raw output from the optical microphone respectively. 
    The third row shows a background noise recording from the optical microphone when no audio signal is played. 
    The fourth row shows the result of subtracting the background noise spectrum in the third row from the recording in the second row. 
	%The second row shows the raw output from the optical microphone.
	%The third row shows the naive technique of subtracting off the noise spectrum from the recorded spectrum.
	The fifth, sixth, and seventh rows show the results of applying the notch filter, the Wiener filter, and both combined, respectively.
	}
\end{figure*}

@@

Results for a selection of filters are collated in Fig.~\ref{fig:BackgroundNotchWienerCombined} where the left and right columns show timeseries and Fourier spectrum results respectively. 
The first and second rows of Fig.~\ref{fig:BackgroundNotchWienerCombined} shows the source signal and the raw optical microphone recording respectively. 
The third row shows a background spectrum of the optical microphone, as shown previously in the top panel Fig.~\ref{fig:psd_noise}.

@@

Given that we have access to the background noise spectrum, an intuitive way to remove noise is via subtracting the noise spectrum from the recorded spectrum. 
The fourth panel of Fig.~\ref{fig:BackgroundNotchWienerCombined} shows the spectrum obtained after subtracting the background noise spectrum. 
We see no obvious improvement, which may be attributed to a time-variant noise spectrum the cause of which is unidentified. %Why the noise sources are time-varying is a fact yet to be determined.

Beyond this na\"{i}ve approach, we apply a range of filters to the data.
In signals processing, the ideal filter would be one that:
%\begin{itemize}
i) completely attenuates the undesired parts of the spectrum, 
ii) does not change the rest of the spectrum, and 
iii) smoothly transitions between these regions, as to not damage the time domain signal when seen under convolution. 
%\end{itemize}
However, these three conditions cannot all hold at once and any filter must compromise between them.

\subsubsection{Comb filter}

Simply zeroing the frequency bins corresponding to the harmonics of the $50\,{\rm Hz}$ mains signal is unsuccessful. This effectively multiplies the spectrum by a rectangular comb filter. It does remove the mains harmonics, but audibly ruins the rest of the signal due to lack of smoothness. This is because applying a filter in frequency space is equivalent to convolving the signal with the inverse Fourier transform of that filter. The inverse Fourier transform of a rectangular comb filter (a set of boxcars) is some combination of sinc functions, which significantly corrupt the signal. 
See also Section~\ref{sec:notch} where we explore a notch filter.

@@

A high-pass filter smoothly attenuates frequencies below some cut-off frequency. Applying a high-pass filter, with a cut-off frequency around $150\,{\rm Hz}$, to the signal spectrum works well at removing the $50\,{\rm Hz}$ and $100\,{\rm Hz}$ harmonics. However, the mains harmonics above $100\,{\rm Hz}$ remain. Using a high-pass filter with a higher cut-off can be used to mitigate this issue. However, it makes the played-back signal unrecognisable as the region above $100\,{\rm Hz}$ carries a lot of the fundamental frequencies of speech and music~\cite{speech_intelligibility}.
Often in speech processing the logarithm of the signal is taken since the amplitude information seems to be more important to intelligibility than the phase information. However, applying a high-pass filter to the logarithm of the signal spectrum does not significantly improve on the above simple high-pass filter.

@@

\begin{figure}
	\includegraphics[width=0.49\textwidth]{figures/butterworth_150_3000.pdf}
	\caption{Butterworth bandpass filter frequency response, any amplitude beyond $-3\,{\rm dB}$ is significant attenuation (half power). The red, dashed vertical lines show the band limits of $150\,{\rm Hz}$ and $3\,{\rm kHz}$. Note the flat response within the band.}
	\label{fig:butterworth}
\end{figure}

A band-pass filter combines a high-pass filter and a low-pass filter to smoothly attenuate frequencies outside of some band. A Butterworth band-pass filter is just a particular choice of the frequency response (an attenuation at each frequency) as to be `maximally flat' within the pass-band region. The Butterworth low-pass component is given by Eqn.~\ref{eq:butterworth} and is combined with a similar high-pass filter to form the band-pass filter.

@@

In the above expression, $f_c$ represents the cut-off frequency of the low-pass Butterworth filter, $\varepsilon$ represents the gain, and $n$ is the order of the filter which determines how quickly the response rolls off past the cut-off frequency. Fig.~\ref{fig:butterworth} shows the frequency response of the filter used here, a fifth order ($n = 5$) Butterworth filter with a pass-band of $(150\,{\rm Hz}$, $3\,{\rm kHz})$. This high frequency cut-off is chosen since the frequencies important for speech generally lie below $2\,{\rm kHz}$~\cite{speech_intelligibility}.

@@

\subsubsection{Cascade notch filter}
\label{sec:notch}
\begin{figure*}
\begin{center}
\includegraphics[width=.9\textwidth]{figures/notch_filter_response.jpg}
\end{center}
\caption{\label{fig:notchMagResponse}
Magnitude response for the first five notches of the $15$ cascaded notch filter described by Eqns.~\ref{eqn:notch} and ~\ref{eqn:notch15}. 
}
\end{figure*}

One method to remove mains noise and its harmonics is to use a sequence of notch filters centred at each of the frequency bins we want to remove. This sequence is known as a cascade of notches. Here the notches are smooth in comparison to the naive zeroing of each frequency (which looks like a rectangular comb). The complex frequency response of a typical infinite impulse response (IIR) notch filter can be written as \citep{10.5555/541204}
\begin{equation}
    \label{eqn:notch}
    H(z)=\frac{1+\alpha}{2}\frac{1-2\beta z^{-1}+z^{-2}}{1-\beta(1+\alpha)z^{-1}+\alpha z^{-2}}
\end{equation}
Where $z$ is the complex frequency and $\alpha$ and $\beta$ are parameters that control the filter. $w_0=\cos^{-1}(\beta)$ is the frequency that is completely attenuated (zeroed) at the centre of the notch and $B_w=\cos^{-1}[2\alpha/(1+\alpha^2)]$ is the bandwidth of the notch (which reflects how quickly the response changes around the notched frequency).
We find that a sequence of $15$ notches works well here, with the $k^\mathrm{th}$ notch centred on the $k^\mathrm{th}$ harmonic of the $50\,{\rm Hz}$ mains signal, where $k=0,2,\dots,14$. When then choose the bandwidth and order (here equal to six) of each notch to avoid disturbing useful signals while still allowing for uncertainty in the location of each harmonic of the mains signal. The response of this cascaded notch filter $H(z)$ is the product of the responses of each of the individual $H_k(z)$ notches,
\begin{equation}
    \label{eqn:notch15}
    H(z) = \prod_{k=0}^{14} H_k(z).
\end{equation}

We use the built-in MATLAB filter design toolbox in this work. 
The amplitude response of the first five notch filters is shown in Fig.~\ref{fig:notchMagResponse}. 
The time series and spectrum obtained after applying the cascade notch filter to the speech recording are shown in the fifth row of Fig.~\ref{fig:BackgroundNotchWienerCombined}.
We see that the mains harmonics are significantly attenuated in comparison background subtracted results shown in the fourth row of Fig.~\ref{fig:BackgroundNotchWienerCombined}.

@@

% This effect will become more severe if the original human spectrum overlaps with the hum spectrum by a large amount.
To overcome this, we turn to more advanced techniques starting with the Wiener filter. Instead of just passively filtering different frequencies, this statistical technique optimises an estimate of the original signal. 
% By purely minimizing the "distance" between estimated signal (filtered human voice) and observed signal (original human voice), we hope that we keep signal part undisturbed while filtering out noise.

@@

\subsubsection{Wiener Filter}

A Wiener filter is an advanced statistical technique that estimates the injected signal given prior information about the injected spectrum and the reference spectrum of the background noise. Let the observed noisy speech signal sequence be given as $\mathbf{x}=\{x(0),\dots, x(N-1)\}$, where $N$ is the length of the data sequence. It is the sum of the original injected signal $\mathbf{s}=\{s(0),\dots,s(N-1)\}$, and the noise sequence $\mathbf{w}=\{w(0),\dots,w(N-1)\}$, 
\begin{equation}
    \mathbf{x}=\mathbf{s}+\mathbf{w}\,.
\end{equation}
Given the observed data $\textbf{x}$, our goal is to make an estimate $\hat{\textbf{s}}$ of the original signal $\textbf{s}$ such that we minimise the Bayesian mean square error (BMSE) between the two, defined as 
\begin{equation}
\label{eq:BMSE}
\text{BMSE}(\hat{\textbf{s}})=E[(\textbf{s}-\hat{\textbf{s}})^2]\,,
\end{equation}

>>>>

% Optical microphone
% aim of optical microphone: play back injected sounds
% some editing in response to pnp
In this section, we explore how this demonstration can be used to teach a selection of signal processing techniques.
The apparatus used in Sections~\ref{sec:ifo} to~\ref{sec:viterbi_wandering} affords the opportunity to explore a range of signal processing techniques and filters of interest to undergraduate students and provide a foundations for many technical applications~\cite{DigitalProcingOfSpeechSignals:1978}. 
We provide a series of example activities using complex audio (such as music and speech) as a natural successor to constant and wandering tones. 
%A natural successor to constant and wandering tones is complex audio, such as music and speech. 
As complex audio signals are not quasi-monochromatic, unlike continuous gravitational wave signals, the Viterbi algorithm as implemented in Section~\ref{sec:viterbi_wandering} is not directly applicable here. Instead, in this section, we use the interferometer as an ``optical microphone'' to detect and play back complex audio signals (i.e.\ using light to capture sound).
In doing so, we experiment with a hierarchy of passive filters which suppress noise, yet make no assumption about the specific form of the signal, unlike the Fourier-based maximum likelihood matched filter which is tuned to the sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood}.
A vast literature exists on signal processing with speech, see Ref.~\cite{DigitalProcingOfSpeechSignals:1978} for introductory material including Fourier analysis and digital representations.
Optical microphone devices have precedence in the laser microphones~\cite{laser_microphone} which are (or were historically) used in the defence industry and operate on a variety of related principles.

%An introduction to ..... can be found in Ref.~\cite{}, including 
%A review of digital processing techniques for speech signals can be found 
%It is also analogous to the detection of unmodelled gravitational wave signals, e.g. broadband bursts from supernovae~\cite{}\han{to do - add citation}. 
%It is also analogous to the detection of unmodelled gravitational wave signals, e.g. broadband bursts from supernovae. [refs; my last sentence can be improved]

This section is indirectly related to the advanced signal processing techniques required to extract gravitational wave signals.
Here, the optical microphone serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories.
We describe the additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and initial results in Section~\ref{sec:initialResultsOpMic}. 
A selection of filter and speech enhancement example activities using the optical microphone are outlined in Section~\ref{sec:filters}.

@@

The human ear can hear frequencies in the range of $\sim 20\,{\rm Hz}$--$20\,{\rm kHz}$. Speech intelligibility (being able to understand speech) requires frequencies up to $3\,{\rm kHz}$ and music requires up to (and beyond) $8\,{\rm kHz}$~\cite{speech_intelligibility}. Therefore, the optical microphone requires a sample rate of at least $16\,{\rm kHz}$ to capture both speech and music (adjusting for the Nyquist frequency). This cannot be achieved with the webcam used in Sections~\ref{sec:single_tone} and ~\ref{sec:viterbi_wandering} as it has a sampling rate of $30\,{\rm Hz}$ and thus can only ``hear'' frequencies below $15\,{\rm Hz}$.
To overcome this issue, we use a photodiode\footnote{A photodiode is an electrical component that acts as a regular diode when no light is incident on it, blocking any current flow in the reverse direction. As the intensity of incident light rises, it becomes increasingly conductive in the reverse direction.} at the output of the interferometer to achieve a sampling rate of $16\,{\rm kHz}$.

We place a OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp which together make a photo-detector that produces a voltage dependent on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-centre position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}, again chosen arbitrarily.
The photodiode is mounted on a cloth screen re-purposed from the dismantled commercial speaker, with the electrical leads connected underneath the screen. 
The voltage signal from the photo-detector is captured by a MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2~\cite{RaspberryPi:online} (henceforth: Pi), which provides a convenient means to record the photodiode data.
Together, the circuit samples the signal at $\sim 16\,{\rm kHz}$. Resources for using the Pi and photodiode are described in Appendix~\ref{app:circuit_diagram}.

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing\footnote{Folding of frequencies greater than half the sampling rate.} into the detected range. We include an anti-aliasing Sallen-Key filter~\cite{sallen_key_filter} tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$, before they are digitally sampled. We also place a second cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

\begin{figure*}
	\includegraphics[width=0.44\textwidth]{figures/ifo_schematic_photodiode_edit.pdf}
	\includegraphics[width=0.52\textwidth]{figures/setup_pic2.pdf}
	\caption{
Left: the updated Michelson interferometer schematic. 
The schematic is identical to Figure~\ref{fig:interference_pattern} except that the data is now recorded using a photodiode and Raspberry Pi instead of a webcam. 
Right: a photograph of the optical microphone. 
The Michelson interferometer is shown on the left and the circuit and Raspberry Pi on the right. 
In the main image, the photodiode is placed behind the cloth screen. 
The inset shows a face-on view of the photodiode with the covering screen removed. 
See Appendix~\ref{app:circuit_diagram} for a circuit diagram and detailed photograph of the circuit.
}
	\label{fig:ifo_schematic_podo}
\end{figure*}

@@

%The source audio was played through the speaker with the optical microphone recording for the full duration of the audio (around a minute long for most sources). 
We test the optical microphone with a variety of recordings, including speech from different people and music ranging from simple melodies and rhythms to songs. During recordings, care is taken to minimise activity around the demonstration to reduce the amount of environmental noise coupling into the interferometer. The time series data is then directly converted to a .wav file and played as an audio recording using the \texttt{scipy.io.wavfile.write} function in Python~\cite{scipy,python}. When processing the results, we restrict our analysis to only the first $10\,{\rm s}$ of each observation (for efficiency), and only plot the first second here.

@@

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/psd_butterworth_14_6.pdf}
	\caption{\label{fig:psd_noise}
Power spectral density (PSD) of background noise (top panel) from the optical microphone (with the speaker off), and the PSD after applying a Butterworth bandpass filter (bottom panel) between the two frequencies marked with red, dashed lines. 
We see strong power from the $50\,{\rm Hz}$ mains hum and its harmonics (most likely from the photodiode circuit and the room’s lighting and cooling). Otherwise, the PSD is fairly white. 
After filtering we see strong attenuation (at least 3~dB) of all frequencies outside the band, but little change to the harmonics within the band.
}
\end{figure}

@@

Here, we describe several filters used to remove the $50\,{\rm Hz}$ mains hum and its harmonics, but also to ultimately improve the speech intelligibility of the recording (to be able to understand the words spoken). We start with some naive approaches and then move to traditional signal processing filters like band-passing and cascaded notches. We finish by combining these with advanced statistical techniques like a Wiener filter and the logMMSE estimator. As an illustrative example, we test all of these filters on the same $1\,{\rm s}$ long speech recording.

\begin{figure*}
	\begin{center}
    %% trim order is <left> <lower> <right> <upper>
    % height=0.5\textheight,trim={1cm 4.5cm 0.4cm 3.5cm},clip
	\includegraphics[width=.8\textwidth, trim={1cm 3cm 1cm 1cm}]{figures/notch_and_wiener_superplot_v2.pdf}
	\end{center}
	\caption{\label{fig:BackgroundNotchWienerCombined}
	Time series (left column) and frequency spectrum (right column) results for the background subtraction, notch, and Wiener filters.
	The first and second rows show the original speech recording and raw output from the optical microphone, respectively. 
    The third row shows a background noise recording from the optical microphone when no audio signal is played. 
    The fourth row shows the result of subtracting the background noise spectrum in the third row from the recording in the second row. 
	%The second row shows the raw output from the optical microphone.
	%The third row shows the naive technique of subtracting off the noise spectrum from the recorded spectrum.
	The fifth, sixth, and seventh rows show the results of applying the notch filter, the Wiener filter, and both combined, respectively.
	}
\end{figure*}

@@

Results for a selection of filters are collated in Fig.~\ref{fig:BackgroundNotchWienerCombined} where the left and right columns show timeseries and Fourier spectrum results, respectively. 
The first and second rows of Fig.~\ref{fig:BackgroundNotchWienerCombined} show the source signal and the raw optical microphone recording, respectively. 
The third row shows a background spectrum of the optical microphone, as shown previously in the top panel Fig.~\ref{fig:psd_noise}.

@@

Given that we have access to the background noise spectrum, an intuitive way to remove noise is via subtracting the noise spectrum from the recorded spectrum. 
The fourth panel of Fig.~\ref{fig:BackgroundNotchWienerCombined} shows the spectrum obtained after subtracting the background noise spectrum. 
We see no obvious improvement, which may be attributed to a time-variant noise spectrum, the cause of which is unidentified. %Why the noise sources are time-varying is a fact yet to be determined.

Beyond this naive approach, we apply a range of filters to the data.
% \jam{changed na\"{i}ve to naive, if you change this back please do it globally}
In signals processing, the ideal filter would be one that:
%\begin{itemize}
i) completely attenuates the undesired parts of the spectrum, 
ii) does not change the rest of the spectrum, and 
iii) smoothly transitions between these regions, as to not damage the time domain signal when seen under convolution. 
%\end{itemize}
However, these three conditions cannot all hold at once and any filter must compromise between them, as we will see below.

\subsubsection{Rectangular comb filter}

Simply zeroing the frequency bins corresponding to the harmonics of the $50\,{\rm Hz}$ mains signal is unsuccessful. This effectively multiplies the spectrum by a rectangular comb filter. It does remove the mains harmonics, but audibly ruins the rest of the signal due to lack of smoothness. This is because applying a filter in frequency space is equivalent to convolving the time-domain signal with the inverse Fourier transform of that filter. The inverse Fourier transform of a rectangular comb filter (a set of boxcars) is some combination of sinc functions, which significantly corrupt the signal. 
See also Section~\ref{sec:notch} where we explore a notch filter.

@@

A high-pass filter smoothly attenuates frequencies below some cut-off frequency. Applying a high-pass filter, with a cut-off frequency around $150\,{\rm Hz}$, to the signal spectrum works well at removing the $50\,{\rm Hz}$ and $100\,{\rm Hz}$ harmonics. However, the mains harmonics above $100\,{\rm Hz}$ remain. Using a high-pass filter with a higher cut-off can be used to mitigate this issue. However, it makes the played-back signal unrecognisable as the region above $100\,{\rm Hz}$ carries a lot of the fundamental frequencies of speech and music~\cite{speech_intelligibility}.
Often in speech processing, the logarithm of the signal is taken since the amplitude information seems to be more important to intelligibility than the phase information to the human ear~\cite{SubjectiveComparison}. However, applying a high-pass filter to the logarithm of the signal spectrum does not significantly improve on the above simple high-pass filter.

@@

\begin{figure}
	\includegraphics[width=0.49\textwidth]{figures/butterworth_150_3000.pdf}
	\caption{Butterworth bandpass filter frequency response. Any amplitude beyond $-3\,{\rm dB}$ is significant attenuation (half power). The red, dashed vertical lines show the band limits of $150\,{\rm Hz}$ and $3\,{\rm kHz}$. Note the flat response within the band characteristic of the Butterworth filter.}
	\label{fig:butterworth}
\end{figure}

A general band-pass filter combines a high-pass filter and a low-pass filter to smoothly attenuate frequencies outside of some band (alt.\ pass-band). A Butterworth band-pass filter is a particular band-pass filter such that the frequency response (the attenuation at each frequency) is ``maximally flat'' within the band. The Butterworth low-pass component is given by Eqn.~\ref{eq:butterworth} and is combined with a similar high-pass filter to form the band-pass filter.

@@

In the above expression, $f_c$ is the cut-off frequency of the low-pass Butterworth filter, $\varepsilon$ is the gain, and $n$ is the order of the filter which determines how quickly the response rolls off past the cut-off frequency. Fig.~\ref{fig:butterworth} shows the frequency response of the filter used here, a fifth order ($n = 5$) Butterworth filter with a pass-band of $(150\,{\rm Hz}$, $3\,{\rm kHz})$. This high frequency cut-off is chosen since the frequencies important for speech generally lie below $2\,{\rm kHz}$~\cite{speech_intelligibility}.

@@

\subsubsection{Cascade notch filter}
\label{sec:notch}
\begin{figure*}
\begin{center}
\includegraphics[width=.9\textwidth]{figures/notch_filter_response.jpg}
\end{center}
\caption{\label{fig:notchMagResponse}
Amplitude (magnitude) response for the first five notches of fifteen for the cascaded notch filter described by Eqs.~\ref{eqn:notch} and~\ref{eqn:notch15}. 
}
\end{figure*}

One method to remove mains noise and its harmonics is to use a sequence of notch filters centred at each of the frequency bins we want to remove. This sequence is known as a ``cascade'' of notches. Here, the notches are smooth in comparison to the naive zeroing of each frequency (which looks like a rectangular comb). The complex frequency response of a typical infinite impulse response (IIR) notch filter can be written as \citep{10.5555/541204}
\begin{equation}
    \label{eqn:notch}
    H(z)=\frac{1+\alpha}{2}\frac{1-2\beta z^{-1}+z^{-2}}{1-\beta(1+\alpha)z^{-1}+\alpha z^{-2}}
\end{equation}
Where $z$ is the complex frequency and $\alpha$ and $\beta$ are parameters that control the filter. $w_0=\cos^{-1}(\beta)$ is the frequency that is completely attenuated (zeroed or ``notched'') at the centre of the notch and $B_w=\cos^{-1}[2\alpha/(1+\alpha^2)]$ is the bandwidth of the notch (which determines how quickly the response changes around the notched frequency).
We find that a sequence of $15$ notches works well here, with the $k^\mathrm{th}$ notch centred on the $k^\mathrm{th}$ harmonic of the $50\,{\rm Hz}$ mains signal, where $k=0,2,\dots,14$. When then choose the bandwidth and order (here equal to six) of each notch to avoid disturbing useful signals while still allowing for uncertainty in the location of each harmonic of the mains signal. The response of this cascaded notch filter $H(z)$ is the product of the responses of each of the individual $H_k(z)$ notches,
\begin{equation}
    \label{eqn:notch15}
    H(z) = \prod_{k=0}^{14} H_k(z).
\end{equation}

We use the built-in MATLAB filter design toolbox in this work~\cite{MATLAB}. 
The amplitude (magnitude) response of the first five notch filters is shown in Fig.~\ref{fig:notchMagResponse}. 
The time series and spectrum obtained after applying the cascade notch filter to the speech recording are shown in the fifth row of Fig.~\ref{fig:BackgroundNotchWienerCombined}.
We see that the mains harmonics are significantly attenuated in comparison background subtracted results shown in the fourth row of Fig.~\ref{fig:BackgroundNotchWienerCombined}.

@@

% This effect will become more severe if the original human spectrum overlaps with the hum spectrum by a large amount.
To overcome this, we turn to more advanced techniques, starting with the Wiener filter. Instead of just passively filtering different frequencies, this statistical technique optimises an estimate of the original signal. 
% By purely minimizing the "distance" between estimated signal (filtered human voice) and observed signal (original human voice), we hope that we keep signal part undisturbed while filtering out noise.

@@

\subsubsection{Wiener filter}
\label{sec:Wiener}

A Wiener filter is an advanced statistical technique that estimates the injected signal given prior information about the injected spectrum and the reference spectrum of the background noise. Let the observed noisy speech signal sequence be given as $\mathbf{x}=(x(0),\dots, x(N-1))$, where $N$ is the length of the data sequence. It is the sum of the original injected signal $\mathbf{s}=(s(0),\dots,s(N-1))$, and the noise sequence $\mathbf{w}=(w(0),\dots,w(N-1))$, 
\begin{equation}
    \mathbf{x}=\mathbf{s}+\mathbf{w}\,.
\end{equation}
Given the observed data $\textbf{x}$, our goal is to make an estimate $\hat{\textbf{s}}$ of the original signal $\textbf{s}$ such that we minimise the Bayesian mean-square-error (BMSE) between the two, defined as 
\begin{equation}
\label{eq:BMSE}
\text{BMSE}(\hat{\textbf{s}})=E[(\textbf{s}-\hat{\textbf{s}})^2]\,,
\end{equation}
